---
title: "March 8, Multilevel II"
author: "Stephen R. Proulx"
date: "3/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
source("../helper.R")
```

# Today's objectives:  
* Simulate data to observe shrinkage (aka partial pooling)
* Apply a partial pooling model to data with a predictor


## Simulating to show an advantage of partial pooling

Define the parameters to use in this simulation
```{r}
## R code 13.7
a_bar <- 1.5
sigma <- 1.5
nponds <- 60
```
```{r}
inv_logit(1.5)
```



Simulate the data
```{r}
set.seed(5005)
sim_data <- tibble(Ni = rep(c(5,10,25,35),each=15),
                   a_pond = rnorm( nponds , mean=a_bar , sd=sigma )) %>%
  mutate(
    p_true = inv_logit(a_pond),
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni)) %>%
  rowid_to_column("pond")


```

Calulate the full-pooling and no pooling maximum likelihood estimates (i.e. maximum posterior values with a flat prior). These are just the global average survival (full-pooling) and the survival within each pond (no-pooling).
```{r}
sim_data <- mutate(sim_data,
                   p_nopool=Si/Ni,
                   p_fullpool=sum(Si)/sum(Ni) )
```


Run the mutli-level model.
```{r}
## R code 13.13
dat <- select( sim_data, pond, Ni, Si )
m13.3 <- ulam(
    alist(
        Si ~ dbinom( Ni , p ),
        logit(p) <- a_pond[pond],
        a_pond[pond] ~ dnorm( a_bar , sigma ),
        a_bar ~ dnorm( 0 , 1.5 ),
        sigma ~ dexp( 1 )
    ), data=dat , chains=4 ,cores=4)
```

```{r}
## R code 13.14
precis( m13.3 , depth=2 )

```

Extract the samples from the model, convert them to the probability scale, and calculate the error from true

```{r}
## R code 13.15
post <- extract.samples( m13.3 ) %>% as_tibble()


# get the a_pond portion of the posterior and summarize
post_a_pond <- select(post,a_pond)%>%
  as.matrix()%>%  as_tibble() %>%
  gather( "ponds","a_pond",1:nponds)%>%
  separate(ponds,c("name","pond"),sep=7) %>%
  mutate(pond=as.numeric(pond),
         p_pond = inv_logit(a_pond)) %>%
  select(-name)

post_a_pond_summarized <- group_by(post_a_pond,pond) %>%
  summarise(
    p_partpool = mean(p_pond),
  )%>%
  ungroup()



# get the a_bar portion of the posterior and summarize
post_a_bar <- select(post,a_bar) 

post_a_bar_summarized <- summarise(post_a_bar,
    a_bar = mean(a_bar))


# put them all in one dataframe
sim_data <- left_join(sim_data,post_a_pond_summarized) %>%
  mutate(partpool_a_bar = post_a_bar_summarized$a_bar[1],
         nopool_p_bar = sum(p_nopool)/nponds,
         nopool_a_bar = logit(nopool_p_bar),
         true_a_bar = sum(a_pond)/nponds
         ) %>%
  select(-nopool_p_bar)

```

Take a minute to look at the error.
```{r}
view(sim_data)
```


compute the error measurement and plot. no pool model fit is in blue.
```{r}
## R code 13.17
sim_data <- sim_data %>%
  mutate(nopool_error = abs(p_nopool-p_true), partpool_error = abs(p_partpool-p_true))


dfline <- 
  sim_data %>%
  select(Ni, nopool_error:partpool_error) %>%
  gather(key, value, -Ni) %>%
  group_by(key, Ni) %>%
  summarise(mean_error = mean(value)) %>%
  mutate(x    = c( 1, 16, 31, 46),
         xend = c(15, 30, 45, 60))


 ggplot(sim_data,aes(x = pond)) +
  geom_vline(xintercept = c(15.5, 30.5, 45.4), 
             color = "white", size = 2/3) +
  geom_point(aes(y = nopool_error), color = "blue") +
  geom_point(aes(y = partpool_error), shape = 1) +
  geom_segment(data = dfline, 
               aes(x = x, xend = xend, 
                   y = mean_error, yend = mean_error),
               color = rep(c("blue", "black"), each = 4),
               linetype = rep(1:2, each = 4)) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 60)) +
  annotate("text", x = c(15 - 7.5, 30 - 7.5, 45 - 7.5, 60 - 7.5), y = .45, 
           label = c("tiny (5)", "small (10)", "medium (25)", "large (35)")) +
  labs(y        = "absolute error",
       x = "pond number")+
  theme(panel.grid    = element_blank(),
        plot.subtitle = element_text(size = 10))
```

### re-simulate and look again
```{r}
sim_data <- tibble(Ni = rep(c(5,10,25,35),each=15),
                   a_pond = rnorm( nponds , mean=a_bar , sd=sigma )) %>%
  mutate(
    p_true = inv_logit(a_pond),
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni)) %>%
  rowid_to_column("pond")


sim_data <- mutate(sim_data,
                   p_nopool=Si/Ni,
                   p_fullpool=sum(Si)/sum(Ni) )


dat <- select( sim_data, pond, Ni, Si )
m13.3new <- ulam(
    alist(
        Si ~ dbinom( Ni , p ),
        logit(p) <- a_pond[pond],
        a_pond[pond] ~ dnorm( a_bar , sigma ),
        a_bar ~ dnorm( 0 , 1.5 ),
        sigma ~ dexp( 1 )
    ), data=dat , chains=4 ,cores=4)
```
```{r}
precis(m13.3new,depth=2)
```


```{r}
## R code 13.15
post <- extract.samples( m13.3new )%>% as_tibble()


# get the a_pond portion of the posterior and summarize
post_a_pond <- select(post,a_pond)%>%
  as.matrix()%>%  as_tibble() %>%
  gather( "ponds","a_pond",1:nponds)%>%
  separate(ponds,c("name","pond"),sep=7) %>%
  mutate(pond=as.numeric(pond),
         p_pond = inv_logit(a_pond)) %>%
  select(-name)

post_a_pond_summarized <- group_by(post_a_pond,pond) %>%
  summarise(
    p_partpool = mean(p_pond),
  )%>%
  ungroup()



# get the a_bar portion of the posterior and summarize
post_a_bar <- select(post,a_bar) 

post_a_bar_summarized <- summarise(post_a_bar,
    a_bar = mean(a_bar))


# put them all in one dataframe
sim_data <- left_join(sim_data,post_a_pond_summarized) %>%
  mutate(partpool_a_bar = post_a_bar_summarized$a_bar[1],
         nopool_p_bar = sum(p_nopool)/nponds,
         nopool_a_bar = logit(nopool_p_bar),
         true_a_bar = sum(a_pond)/nponds
         ) %>%
  select(-nopool_p_bar)



## R code 13.17
sim_data <- sim_data %>%
  mutate(nopool_error = abs(p_nopool-p_true), partpool_error = abs(p_partpool-p_true))


dfline <- 
  sim_data %>%
  select(Ni, nopool_error:partpool_error) %>%
  gather(key, value, -Ni) %>%
  group_by(key, Ni) %>%
  summarise(mean_error = mean(value)) %>%
  mutate(x    = c( 1, 16, 31, 46),
         xend = c(15, 30, 45, 60))


 ggplot(sim_data,aes(x = pond)) +
  geom_vline(xintercept = c(15.5, 30.5, 45.4), 
             color = "white", size = 2/3) +
  geom_point(aes(y = nopool_error), color = "blue") +
  geom_point(aes(y = partpool_error), shape = 1) +
  geom_segment(data = dfline, 
               aes(x = x, xend = xend, 
                   y = mean_error, yend = mean_error),
               color = rep(c("blue", "black"), each = 4),
               linetype = rep(1:2, each = 4)) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 60)) +
  annotate("text", x = c(15 - 7.5, 30 - 7.5, 45 - 7.5, 60 - 7.5), y = .45, 
           label = c("tiny (5)", "small (10)", "medium (25)", "large (35)")) +
  labs(y        = "absolute error",
       x = "pond number")+
  theme(panel.grid    = element_blank(),
        plot.subtitle = element_text(size = 10))
```

`
```{r}
view(sim_data)
```

### error in the full estimate of a?
```{r , eval=FALSE}
abar_error<-tibble(nopool_error=rep(0,20),partpool_error=rep(0,20))


for(i in 1:20){
  sim_data <- tibble(Ni = rep(c(5,10,25,35),each=15),
                   a_pond = rnorm( nponds , mean=a_bar , sd=sigma )) %>%
  mutate(
    p_true = inv_logit(a_pond),
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni)) %>%
  rowid_to_column("pond")


sim_data <- mutate(sim_data,
                   p_nopool=Si/Ni,
                   p_fullpool=sum(Si)/sum(Ni) )



m13.3new <- stan( fit=m13.3@stanfit , data=sim_data , chains=4 )



## R code 13.15
post <- extract.samples( m13.3new ) %>% as_tibble()


# get the a_pond portion of the posterior and summarize
post_a_pond <- select(post,a_pond)%>%
  as.matrix()%>%  as_tibble() %>%
  gather( "ponds","a_pond",1:nponds)%>%
  separate(ponds,c("name","pond"),sep=7) %>%
  mutate(pond=as.numeric(pond),
         p_pond = inv_logit(a_pond)) %>%
  select(-name)

post_a_pond_summarized <- group_by(post_a_pond,pond) %>%
  summarise(
    p_partpool = mean(p_pond),
  )%>%
  ungroup()



# get the a_bar portion of the posterior and summarize
post_a_bar <- select(post,a_bar) 

post_a_bar_summarized <- summarise(post_a_bar,
    a_bar = mean(a_bar))


# put them all in one dataframe
sim_data <- left_join(sim_data,post_a_pond_summarized) %>%
  mutate(partpool_a_bar = post_a_bar_summarized$a_bar[1],
         nopool_p_bar = sum(p_nopool)/nponds,
         nopool_a_bar = logit(nopool_p_bar),
         true_a_bar = sum(a_pond)/nponds
         ) %>%
  select(-nopool_p_bar)


  
abar_error$partpool_error[i]<-abs(sim_data$partpool_a_bar[1]-sim_data$true_a_bar[1])

abar_error$nopool_error[i]<-abs(sim_data$nopool_a_bar[1]-sim_data$true_a_bar[1])

}
```

(Pre-run data from the above code)
```{r}
abar_error <- tibble(nopool_error=c( 0.4182332, 0.5081091, 0.3939369, 0.3559052, 0.5133194, 0.2753575, 0.4588049, 0.4695861, 0.3774278, 0.6692748, 0.4374436, 0.316275, 0.2648031, 0.5272011, 0.2222871, 0.4567554, 0.2917637, 0.4006492, 0.4792763, 0.2956382),
                     partpool_error=c(0.0683,	0.091,	0.0524,	0.0545,	0.221,	0.0431,	0.125,	0.0289,	0.0111,	0.296,	0.0171,	0.0144,	0.0876,	0.206,	0.102,	0.00177,	0.0449,	0.0276,	0.0899,	0.0391))
```


```{r} 
ggplot(abar_error,aes(x=nopool_error)) +
         geom_histogram(bins=15,fill="blue",alpha=0.3) +
         geom_histogram(aes(x=partpool_error), bins=15,fill="red",alpha=0.3) 
```





## Partial pooling with categories 


Now we will do a simulation with two "treatments". The first set of data points have a mean parameter of `a_bar1` and the second set have a parameter of `a_bar2`. We make two data frames and then put them together. 

These have a fairly different mean value of `a_bar`.
```{r}
a_bar1 <- 1.5
a_bar2 <-0
sigma <- 0.5
nponds <- 50

set.seed(5005)

sim_data1 <- tibble(treatment = 1,Ni = rep(c(50),each=50),
                   a_pond = rnorm( nponds , mean=a_bar1 , sd=sigma )) %>%
  mutate(
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni))

sim_data2 <- tibble(treatment = 2,Ni = rep(c(50),each=50),
                   a_pond = rnorm( nponds , mean=a_bar2 , sd=sigma )) %>%
  mutate(
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni))



sim_data_tot=bind_rows(sim_data1,sim_data2) %>% 
  rowid_to_column("pond")
```


Run the mutlilevel model. We specify a_pond[pond] as before, but now the distribution that it is drawn from has an a_bar that depends on treatment.
```{r}
dat=select(sim_data_tot,-a_pond)

m13.3Abig <- ulam(
    alist(
        Si ~ dbinom( Ni , p ),
        logit(p) <- a_pond[pond],
        a_pond[pond] ~ dnorm( a_bar[treatment] , sigma ),
        a_bar[treatment] ~ dnorm( 0 , 1.5 ),
        sigma ~ dexp( 1 )
    ), data=select(sim_data_tot,-a_pond) , chains=4 , cores=4 ,iter=5000)
```

The precis includes the values for each pond, but go to the end and look at the treatment specific values.
```{r}
precis(m13.3Abig, depth = 2)

```


To see what the effect of treatment is, we need to calculate the contrast for each sample and average over all samples.
```{r}
post <- extract.samples( m13.3Abig ) %>% as_tibble() %>%
  select(a_bar)


contrast<- mutate(post,delta_a=a_bar[,2]-a_bar[,1])


quantile(contrast$delta_a,c(0.05,.5,.95))
```





Alternative version of the model, taking advantage of the additive way that nested normal distributions can be simplified.
```{r}
dat=select(sim_data_tot,-a_pond)

m13.3AbigA <- ulam(
    alist(
        Si ~ dbinom( Ni , p ),
        logit(p) <- a_bar[treatment] +a_pond[pond],
        a_pond[pond] ~ dnorm( 0 , sigma ),
        a_bar[treatment] ~ dnorm( 0 , 1.5 ),
        sigma ~ dexp( 1 )
    ), data=select(sim_data_tot,-a_pond) , chains=4 ,cores=4, iter=5000)
```

```{r}
precis(m13.3Abig, depth = 2)

precis(m13.3AbigA, depth = 2)
```

```{r}
post <- extract.samples( m13.3AbigA ) %>% as_tibble() %>%
  select(a_bar)


contrast<- mutate(post,delta_a=a_bar[,2]-a_bar[,1])


quantile(contrast$delta_a,c(0.05,.5,.95))
```


What happens if the effect is smaller?
```{r}
a_bar1 <- 1.5
a_bar2 <- 1.0
sigma <- 0.5
nponds <- 50

set.seed(5005)

sim_data1 <- tibble(treatment = 1,Ni = rep(c(50),each=50),
                   a_pond = rnorm( nponds , mean=a_bar1 , sd=sigma )) %>%
  mutate(
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni))

sim_data2 <- tibble(treatment = 2,Ni = rep(c(50),each=50),
                   a_pond = rnorm( nponds , mean=a_bar2 , sd=sigma )) %>%
  mutate(
    Si = rbinom(n(),prob=inv_logit(a_pond),size=Ni))



sim_data_tots=bind_rows(sim_data1,sim_data2) %>% 
  rowid_to_column("pond")
```

Re-run the analysis and see how things are different:
